---
title: "3.3 Linear and Quadratic Discrimination"
author: "Ga young Moon"
date: "2025-02-15"
institute: Department of Statistics \newline Sungshin Women’s University
fonttheme: "serif"
fontsize: 8pt
output:
  beamer_presentation:
    latex_engine: xelatex 
    theme: "metropolis"
header-includes:
  - \input{header_includes.tex}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height = 4, fig.width = 6)
set.seed(1)
```

# Linear / Quadratic Discrimination

## Linear / Quadratic Discrimination

In discriminant analysis, the discriminant function is divided into two types according to its form.

In the case of a discriminant function in the form:

-   a straight line $\rightarrow$ Linear Discrimination

-   a quadratic curve $\rightarrow$ Quadratic Discrimination

## Distribution of $x \in \mathbb{R}^p$ given $y=\pm 1$

The distribution of $x \in \mathbb{R}^p$ given $y=\pm 1$

$\rightarrow$ $N(\mu_{\pm1}, \sum_{\pm1})$,

that is, it is assumed to follow a multivariate normal distribution.

$\rightarrow$ Multivariate Normal Distribution: $\newline$ distribution that extends the normal distribution to multidimensional spaces

-   $\mu_{\pm1}$ : means of each features, $\sum_{\pm1}$ : matrix of variance-covariance,

$$
\sum= \: \begin{pmatrix} {\sigma_{11}^2} & \cdots & {\sigma_{1p}^2} \\ \vdots & \ddots & \vdots \\ \sigma_{p1}^2 & \cdots & \sigma_{pp}^2 \end{pmatrix}
$$

## Probability Density Function of $x \in \mathbb{R}^p$ given $y=\pm 1$

Probability Density Function of $x \in \mathbb{R}^p$ given $y = \pm1$:

$$
f_{\pm1}(x) =  \frac{1}{\sqrt{(2\pi)^p
\begin{vmatrix} \sum \end{vmatrix}}} \exp \left\{-\frac{1}{2}(x-\mu_{\pm1})^T\sum^{-1}_{\pm1}(x-\mu_{\pm1})\right\}
$$

# Discriminant Function

## Posterior probability

Assuming that the probabilities of $y= \pm1$ are known before seeing the covariates $x$,

$\rightarrow$ this is called Prior probability.

-   When the probability function of a random variable $X$ is $f_{\pm1}(x)$ and the prior probability at that time is $\pi_{\pm1}$,

The posterior probability:

$$
\begin{aligned}
P(Y|X) &= \frac{P(X=x|Y=\pm1)P(Y= \pm1)}{P(X=x)} \\
&= \frac{\pi_{\pm1} f_{\pm1}(x)}{\pi_1 f_1(x) +\pi_{-1} f_{-1}(x)}
\end{aligned}
$$

## Minimizing the erroe probability

Assuming that:

1.  $f_{\pm1}$ follows a Gaussian distribution,
2.  The expectation $\mu_{\pm1}$ and the covariance matrix $\sum_{\pm1}$ are known,
3.  $\pi_{\pm1}$ is also known,

$$
\frac{\pi_1 f_1(x)}{\pi_1 f_1(x) + \pi_{-1} f_{-1}(x)} \; \geq \; \frac{\pi_{-1} f_{-1}(x)}{\pi_1 f_1(x) + \pi_{-1} f_{-1}(x)}
$$

That is, we can minimize the error probability:

-   by estimating $y=1$, when $\pi_1 f_1(x) \; \geq \; \pi_{-1} f_{-1}(x)$.

-   by estimating $y=-1$, when $\pi_1 f_1(x) \; \leq \; \pi_{-1} f_{-1}(x)$.

## Maximizing the posterior probability

When the number of values is $K$:

Maximizing the posterior probability applies,

not only to the case of $K=2$ but also to the case of $K\geq2$.

-   The probability that $y=k$ is $P(y=k|x)$, when $k= 1, \cdots, K$ and the covariates $x$ are given.

## Maximizing the posterior probability

When estimating $y=\hat k$, the probability that the estimate is correct:

$$
1- \sum_{k \neq \hat k} P(y=k|x) \; = \; 1- P(y= \hat k|x)
$$

-   when the prior probability is known, $\hat k$ minimizes the average error probability.

-   We choose a $k$ that maximizes the posterior probability $P(y= \hat k|x)$.

## The border

For simplicity, let's assume $K=2$, and when maximizing the posterior probability,

$\rightarrow$ we see the properties at the border between $y = \pm1$.

Property:

$$
\scriptstyle 
-(x- \mu_1)^T \sum_1^{-1}(x- \mu_1) + (x- \mu_{-1})^T \sum_{-1}^{-1}(x- \mu_{-1}) \; = \; \log \frac{\begin{vmatrix} \sum_1 \end{vmatrix}}{\begin{vmatrix} \sum_{-1} \end{vmatrix}} - 2\log \frac{\pi_1}{\pi_{-1}}
$$

## The border

In general, the border is

$\rightarrow$ a functoin of the quadratic forms of $x^T \sum_1^{-1}x$ and $x^T \sum_{-1}^{-1} x$

and this case is Quadratic Discrimination.

-   In particular, when $\sum_1 \; = \; \sum_{-1}$ (if we write them as $\sum$), the border becomes a surface , that we call Linear Discrimination.

## The border

When $x^T \sum_1^{-1}x = x^T \sum_{-1}^{-1} x$,

The border becomes:

$$
2(\mu_1 - \mu_{-1})^T \sum^{-1}x - (\mu_1^T \sum^{-1} \mu_1 - \mu_{-1} \sum^{-1} \mu_{-1}) \; = \; -2 \log \frac{\pi_1}{\pi_{-1}}
$$

or more simply,

$$
(\mu_1-\mu_{-1})^T \sum^{-1} (x- \frac{\mu_1 + \mu_{-1}}{2}) \; = \; - \log \frac{\pi_1}{\pi_{-1}}
$$

## The border

If $\pi_1 = \pi_{-1}$,

then the border is $x= \frac{\mu_1 + \mu_{-1}}{2}$.

-   If $\pi_{\pm1}$ and $f_{\pm1}$ are unknown, we need to estimate them from the training data.

# Example 35, 36

## [Example 35] Output the border of the Quadratic Discriminant using the R code

The following code draws the border for estimating the mean and covariance of the covariates $x$ for $y=\pm1$.

```{r code1, echo = TRUE, eval= FALSE}
mu.1=c(2,2); sigma.1=2; sigma.2=2; rho.1=0
mu.2=c(-3,-3); sigma.3=1; sigma.4=1; rho.2=-0.8
n=100
u=rnorm(n); v=rnorm(n); x.1=sigma.1*u+mu.1[1];
y.1=(rho.1*u+sqrt(1-rho.1^2)*v)*sigma.2+mu.1[2]
u=rnorm(n); v=rnorm(n); x.2=sigma.3*u+mu.2[1];
y.2=(rho.2*u+sqrt(1-rho.2^2)*v)*sigma.4+mu.2[2]
f=function(x,mu,inv,de)drop(-0.5*t(x-mu)%*%inv%*%(x-mu)-0.5*log(de))
mu.1=mean(c(x.1,y.1)); mu.2=mean(c(x.2,y.2));
df=data.frame(x.1,y.1); mat=cov(df); inv.1=solve(mat); de.1=det(mat)
df=data.frame(x.2,y.2); mat=cov(df); inv.2=solve(mat); de.2=det(mat)
f.1=function(u,v)f(c(u,v),mu.1,inv.1,de.1);
f.2=function(u,v)f(c(u,v),mu.2,inv.2,de.2)
pi.1=0.5; pi.2=0.5
u = v = seq(-6, 6, length=50); m=length(u); w=array(dim=c(m,m))
for(i in 1:m)for(j in 1:m)w[i,j]=log(pi.1)+f.1(u[i],v[j])-log(pi.2)-f.2(u[i],v[j])
# plot
contour(u,v,w,level=0)
points(x.1,y.1,col="red"); points(x.2,y.2,col="blue")
```

## [Example 35] Output the border of the Quadratic Discriminant using the R code

```{r plot1, echo = FALSE, fig.height= 3.4, fig.width= 4.3, fig.align= 'center'}
mu.1=c(2,2); sigma.1=2; sigma.2=2; rho.1=0
mu.2=c(-3,-3); sigma.3=1; sigma.4=1; rho.2=-0.8
n=100
u=rnorm(n); v=rnorm(n); x.1=sigma.1*u+mu.1[1];
y.1=(rho.1*u+sqrt(1-rho.1^2)*v)*sigma.2+mu.1[2]
u=rnorm(n); v=rnorm(n); x.2=sigma.3*u+mu.2[1];
y.2=(rho.2*u+sqrt(1-rho.2^2)*v)*sigma.4+mu.2[2]
f=function(x,mu,inv,de)drop(-0.5*t(x-mu)%*%inv%*%(x-mu)-0.5*log(de))
mu.1=mean(c(x.1,y.1)); mu.2=mean(c(x.2,y.2));
df=data.frame(x.1,y.1); mat=cov(df); inv.1=solve(mat); de.1=det(mat)
df=data.frame(x.2,y.2); mat=cov(df); inv.2=solve(mat); de.2=det(mat)
f.1=function(u,v)f(c(u,v),mu.1,inv.1,de.1);
f.2=function(u,v)f(c(u,v),mu.2,inv.2,de.2)
pi.1=0.5; pi.2=0.5
u = v = seq(-6, 6, length=50); m=length(u); w=array(dim=c(m,m))
for(i in 1:m)for(j in 1:m)w[i,j]=log(pi.1)+f.1(u[i],v[j])-log(pi.2)-f.2(u[i],v[j])
# plot
contour(u,v,w,level=0)
points(x.1,y.1,col="red"); points(x.2,y.2,col="blue")
```

## [Example 35] Output the border of the Linear Discriminant using the R code

If the covariance matrices are equal, we can use the following code:

```{r code2, echo = TRUE, eval= FALSE}
mu.1=c(2,2); sigma.1=2; sigma.2=2; rho.1=0
mu.2=c(-3,-3); sigma.3=1; sigma.4=1; rho.2=-0.8
n=100
u=rnorm(n); v=rnorm(n); x.1=sigma.1*u+mu.1[1];
y.1=(rho.1*u+sqrt(1-rho.1^2)*v)*sigma.2+mu.1[2]
u=rnorm(n); v=rnorm(n); x.2=sigma.3*u+mu.2[1];
y.2=(rho.2*u+sqrt(1-rho.2^2)*v)*sigma.4+mu.2[2]
f=function(x,mu,inv,de)drop(-0.5*t(x-mu)%*%inv%*%(x-mu)-0.5*log(de))
mu.1=mean(c(x.1,y.1)); mu.2=mean(c(x.2,y.2));
df=data.frame(c(x.1,y.1)-mu.1, c(x.2,y.2)-mu.2); inv.1=solve(mat); de.1=det(mat)
inv.2=inv.1; de.2=de.1
f.1=function(u,v)f(c(u,v),mu.1,inv.1,de.1);
f.2=function(u,v)f(c(u,v),mu.2,inv.2,de.2)
pi.1=0.5; pi.2=0.5
u = v = seq(-6, 6, length=50); m=length(u); w=array(dim=c(m,m))
for(i in 1:m)for(j in 1:m)w[i,j]=log(pi.1)+f.1(u[i],v[j])-log(pi.2)-f.2(u[i],v[j])
# plot
contour(u,v,w,level=0)
points(x.1,y.1,col="red"); points(x.2,y.2,col="blue")
```

## [Example 35] Output the border of the Linear Discriminant using the R code

```{r plot2, echo = FALSE, fig.height= 3.4, fig.width= 4.3, fig.align= 'center'}
mu.1=c(2,2); sigma.1=2; sigma.2=2; rho.1=0
mu.2=c(-3,-3); sigma.3=1; sigma.4=1; rho.2=-0.8
n=100
u=rnorm(n); v=rnorm(n); x.1=sigma.1*u+mu.1[1];
y.1=(rho.1*u+sqrt(1-rho.1^2)*v)*sigma.2+mu.1[2]
u=rnorm(n); v=rnorm(n); x.2=sigma.3*u+mu.2[1];
y.2=(rho.2*u+sqrt(1-rho.2^2)*v)*sigma.4+mu.2[2]
f=function(x,mu,inv,de)drop(-0.5*t(x-mu)%*%inv%*%(x-mu)-0.5*log(de))
mu.1=mean(c(x.1,y.1)); mu.2=mean(c(x.2,y.2));
df=data.frame(c(x.1,y.1)-mu.1, c(x.2,y.2)-mu.2); inv.1=solve(mat); de.1=det(mat)
inv.2=inv.1; de.2=de.1
f.1=function(u,v)f(c(u,v),mu.1,inv.1,de.1);
f.2=function(u,v)f(c(u,v),mu.2,inv.2,de.2)
pi.1=0.5; pi.2=0.5
u = v = seq(-6, 6, length=50); m=length(u); w=array(dim=c(m,m))
for(i in 1:m)for(j in 1:m)w[i,j]=log(pi.1)+f.1(u[i],v[j])-log(pi.2)-f.2(u[i],v[j])
# plot
contour(u,v,w,level=0)
points(x.1,y.1,col="red"); points(x.2,y.2,col="blue")
```

## [Example 35] Output the Quadratic Discriminant graph using the R code

Through two border pictures,

-   We can see that if covariance matrices are equal, the border is a line, otherwise, it is a quadratic curve.
-   In the linear discrimination, if the prior probabilities and the covariance matrices are equal, then the border is the vertical bisector of the line connecting the centers.

## [Example 36] Iris data classification: using the classifier via Quadratic Discrimination

When the response takes more than two values

$\rightarrow$ we can choose the response with the maximum posterior probability.

Fisher's Iris data to use in Ex.36:

-   It contains four covariates, and the response variable which is the three species containing 50 samples. (N= 150, p= 4)

-   We evaluate it using the test data set that is different from the training data set.

## [Example 36] Iris data classification: using the classifier via Quadratic Discrimination

```{r code3, echo= TRUE, eval= FALSE}
f=function(w,mu,inv,de)-0.5*(w-mu)%*%inv%*%t(w-mu)-0.5*log(de)
df=iris; df[[5]]=c(rep(1,50),rep(2,50),rep(3,50))
n=nrow(df); train=sample(1:n,n/2,replace=FALSE); test=setdiff(1:n,train)
mat=as.matrix(df[train,])
mu=list(); covv=list()
for(j in 1:3){
  x=mat[mat[,5]==j,1:4];
  mu[[j]]=c(mean(x[,1]),mean(x[,2]),mean(x[,3]),mean(x[,4]))
  covv[[j]]=cov(x)
}
g=function(v,j)f(v,mu[[j]],solve(covv[[j]]),det(covv[[j]]))
z=array(dim=n/2)
for(i in test){
  u=as.matrix(df[i,1:4]); a=g(u,1);b=g(u,2); c=g(u,3)
  if(a<b){if(b<c)z[i]=3 else z[i]=2}
  else {if(a<c)z[i]=3 else z[i]=1}
}
table(z[test],df[test,5])
```

## [Example 36] Iris data classification: using the classifier via Quadratic Discrimination

```{r table1, echo= FALSE}
f=function(w,mu,inv,de)-0.5*(w-mu)%*%inv%*%t(w-mu)-0.5*log(de)
df=iris; df[[5]]=c(rep(1,50),rep(2,50),rep(3,50))
n=nrow(df); train=sample(1:n,n/2,replace=FALSE); test=setdiff(1:n,train)
mat=as.matrix(df[train,])
mu=list(); covv=list()
for(j in 1:3){
  x=mat[mat[,5]==j,1:4];
  mu[[j]]=c(mean(x[,1]),mean(x[,2]),mean(x[,3]),mean(x[,4]))
  covv[[j]]=cov(x)
}
g=function(v,j)f(v,mu[[j]],solve(covv[[j]]),det(covv[[j]]))
z=array(dim=n/2)
for(i in test){
  u=as.matrix(df[i,1:4]); a=g(u,1);b=g(u,2); c=g(u,3)
  if(a<b){if(b<c)z[i]=3 else z[i]=2}
  else {if(a<c)z[i]=3 else z[i]=1}
}
table(z[test],df[test,5])
```

-   vertical axis: Values classified using the classifier

-   horizontal axis: Values of test data set

-   Setosa and Virginia can be seen as being well classified accroding to the test data set,

-   on the other hand, in the case of Versicolor,

    there are two values that were incorrectly classified as Virginia when analyzed by the QDA classifier.